### 2019-12-06 reports/article.tex
fixed acknowledgments
article revisions

% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}
% etc.
%
\usepackage[misc]{ifsym} %
\usepackage{nameref,hyperref}
\usepackage{booktabs}
\usepackage{changepage}
% \usepackage{geometry}

@@

\author{Mart\'i Bosch         \and
        J\'er\^ome Chenal %etc.
}

@@

\institute{Mart\'i Bosch (\Letter) \and J\'er\^ome Chenal \at
              Urban and Regional Planning Community (CEAT), \'Ecole Polytechnique F\'ed\'erale de Lausanne (EPFL), Lausanne, Switzerland \\
              Tel.: +41 21 69 34435\\
              % Fax: +123-45-678910\\
              \email{marti.bosch@epfl.ch}           %  \\
%             \emph{Present address:} of F. Author  %  if needed
           \and
           J\'er\^ome Chenal \at
              \email{jerome.chenal@epfl.ch}
}

@@

\begin{abstract}
  Urbanization is currently a global phenomenon that has become the most important form of landscape change and is increasingly affecting biodiversity and ecosystem functions.
In order to evaluate the impacts of urbanization and inform urban planning, it is important to understand the spatiotemporal patterns of land use change associated to urbanization.
This paper exploits three different frameworks, namely landscape metrics, urban growth modes and fractal analysis to characterize the spatiotemporal patterns of urbanization of the Swiss urban agglomerations of Zurich, Bern and Lausanne.
The land use inventory provided by the Swiss Federal Statistical Office was used to assemble four temporal snapshots from 1980 to 2016 at the extent of the urban agglomerations.
The time series of landscape metrics generally supports the diffusion and coalescence model of urban growth, with Zurich exhibiting most characteristics of coalescence while Bern and Lausanne seem to be at the transition between diffusion and coalescence.
Nevertheless, the analysis of the urban growth modes suggest that leapfrog development occurs at all periods, which contributes to an increasing fragmentation of natural patches and maintains the fractal configuration of the landscape.
The discussion reviews potential explanations for the observed landscape changes, and concludes with some planning implications.

@@

The last centuries have seen an unprecedented growth of urban areas, which has resulted in dramatic conversion of natural land and profound changes in landscape patterns and the ecosystem functions that they support \citep{alberti2005effects}.
The combination of current demographic prospects and the observed trends of decreasing urban densities suggest that the global amount of land occupied by cities might increase threefold \citep{angel2005dynamics}.
Quantifying urban landscape patterns in space and time is an important and necessary step to understand the driving forces and ecological impacts of urbanization \citep{wu2014urban}.

Recent decades have witnessed an increasing number of studies of the spatiotemporal patterns of land use change associated to urbanization \citep{dietzel2005spatio, seto2005quantifying, schneider2008compact, jenerette2010global, wu2011quantifying, li2013quantifying, liu2016general, nong2018quantifying}.
Initial attempts to synthesis suggested that urbanization can be characterized as a two-step alternating process of diffusion and coalescence \citep{dietzel2005spatio, schneider2008compact}, nonetheless, subsequent studies challanged the empirical validity of such hypothesis.
On the one hand, \cite{jenerette2010global} examined the spatiotemporal patterns of land use change of a sample of 120 cities distributed throughout the world from 1990 to 2000, and determined that overall, urbanization leads to fragmented landscapes with more complex and heterogeneous structures. On the other hand, \cite{li2013quantifying} determined that the two-phase diffusion and coalescence model can be over-simplistic and that urbanization might be better characterized by means of three growth modes, namely infilling, edge expansion and leapfrogging, which operate simoultaneously while alternating their relative dominance.

Paralleling the above studies, approaches from the complexity sciences have provided novel insights into the spatial organization that underpins contemporary cities \citep{batty2005cities}.
Like many other complex systems, cities exhibit morphological traits that are consistent with fractal geometry and reflect the self-organizing nature of the processes that occur upon them \citep{white1993cellular}.
Although the scaling relationships of fractal geometry suggest the existence of strong morphological regularities over a wide variety of cities and regions \citep{frankhauser1994fractalite, batty1994fractal, white2015modeling}, their meaning in the context of the spatiotemporal patterns of urbanization remains unclear \citep{li2000fractal, manson2006complexity, bosch2019addressing}.

This paper intends to evaluate the spatiotemporal patterns of land use change observed in three of the largest Swiss urban agglomerations from three different perspectives, namely landscape metrics, urban growth modes and fractal analysis. 
The objective of this paper is twofold.
The first is to assess to test whether the spatiotemporal evolution of the three agglomerations conform to the diffusion and coalescence dichotomy, the second is to explore how the three adopted perspectives complement each other.
The results will serve to discuss planning implications regarding the desirability of the recent densification policies adopted in Switzerland.

@@

Given that a significant part of the cross-border urban agglomerations of Geneva and Basel (the second and third largest in Switzerland) lie beyond the Swiss boundaries \mycitep{sfso2014espace}, in order to ensure coherence of the land use/land cover data (see the section below), this study comprises only three of the five largest Swiss urban agglomerations, namely Zurich, Bern and Lausanne \mycitep{sfso2018city}.
As shown in \autoref{figures/population_change.pdf}, the three agglomerations have undergone important population growth over the last 30 years, especially during the most recent years and at the agglomeration extent.
With a total population over 1.3 million and land area of 1305 km$^2$ (1038 hab/km$^2$), Zurich is the largest Swiss urban agglomeration. As a leading global city and one of the world's largest financial centers, Zurich has the country's largest airport and railway station, and also hosts the largest Swiss universities and higher education institutions. 
Bern is the capital of Switzerland and fourth most populous urban agglomeration in Switzerland, with a total population of 410000 inhabitants and occupying a land area of 783 km$^2$ (531 hab/km$^2$). As the fifth largest Swiss urban agglomeration and the second most important student and research center after Zurich, the Lausanne agglomeration has a total population of 409000 inhabitants over a land area of 773 km$^2$ (537 hab/km$^2$). Given its larger population growth rate, Lausanne is likely to soon surpass Bern and become the fourth largest urban agglomeration in Switzerland.
% Zurich: 1021859 (1990), 1080728 (2000), 1191058 (2010), 1354140 (2017)
% Bern: 351084 (1990), 349157 (1990), 357668 (2010), 415785 (2017)
% Lausanne: 300280 (1990), 311441 (2000), 339,389 (2010), 415596 (2017)
% Densities: Zurich: 1037 Bern: 531, Lausanne: 537
Overall, the three urban agglomerations are characterized by a pervasive public transportation system and a highly developed economy, with a 85\% of the employment devoted to the tertiary sector.

@@

\begin{figure}[!ht]
  \begin{adjustwidth}{0cm}{-.4\textwidth}
    \centering  
    \includegraphics[width=\linewidth]{figures/landscape_plots.pdf}
    % \vspace{.5em}
    \caption[Evolution of urban patches]{\label{figures/landscape_plots.pdf}Evolution of urban patches of the three urban agglomerations throughout their respective periods of study. The times t$_0$, t$_1$, t$_2$ and t$_3$ correspond to 1981, 1993, 2004 and 2013 for Bern; 1980, 1990, 2005 and 2014 for Lausanne and 1982, 1994, 2007 and 2016 for Zurich}
  \end{adjustwidth}
\end{figure}

@@

While a plentiful collection of landscape metrics can be find in the literature, many of them are highly correlated with one another. As a matter of fact, \cite{riitters1995factor} found that the characteristics discerned by 55 prevalent landscape metrics could be reduced to only 6 independent factors.
On the other hand, landscape metrics can be very sensitive to the resolution and the extent of the maps. However, several metrics empirically exhibit consistent responses to changing scales that conform to predictable scaling relations \citep{wu2002empirical, wu2004effects}.
Based on such remarks, and in order to enhance comparability with other studies, ten landscape metrics have been selected for the present study, whose details are listed in \autoref{selected-metrics}.
While complying with the FRAGSTATS v4 definitions \citep{mcgarigal2012fragstats}, the landscape metrics have been computed with the open source library PyLandStats \citep{bosch2019pylandstats}.
Like in most of the related studies, the categorical maps have been reclassified into urban and natural classes, and the metrics have computed at the urban class level, namely aggregating their values across all the urban patches of the landscape. The only exceptions are the contagion (CONTAG) and Shannon's diversity index (SHDI), which can only be computed at the landscape level, namely considering all the landscape classes (urban and natural in the this study).
Pixels that correspond to land unavailable for development, such as water bodies, have been excluded from the computation of the metrics.

>>>>

% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}
% etc.
%
\usepackage[misc]{ifsym} %
\usepackage{nameref,hyperref}
\usepackage{booktabs}
\usepackage{changepage}
% \usepackage{geometry}
\usepackage{subcaption}

@@

\author{
  Mart\'i Bosch \and
  R\'emi Jaligot \and
  J\'er\^ome Chenal
}

@@

\institute{
  Mart\'i Bosch (\Letter) \and R\'emi Jaligot \and J\'er\^ome Chenal \at
  Urban and Regional Planning Community (CEAT), \'Ecole Polytechnique F\'ed\'erale de Lausanne (EPFL), Lausanne, Switzerland \\
  Tel.: +41 21 69 34435\\
  % Fax: +123-45-678910\\
  \email{marti.bosch@epfl.ch}           %  \\
  % \emph{Present address:} of F. Author  %  if needed
  \and
  R\'emi Jaligot \at
  \email{remi.jaligot@epfl.ch}  
  \and
  J\'er\^ome Chenal \at
  \email{jerome.chenal@epfl.ch}
}

@@

\begin{abstract}
  \hfill\break

  \emph{Context} Urbanization is the most important form of landscape change and is increasingly affecting biodiversity and ecosystem functions.
  Understanding how landscape patterns change in space and time is central to the evaluation of the environmental impacts of urbanization.

  \emph{Objectives} This research explores the spatiotemporal patterns of land use change in the Swiss urban agglomerations of Bern, Lausanne and Zurich at two characteristic spatial extents, and compares them to prominent hypotheses of urbanization patterns. % The discussion reviews potential explanations for the observed landscape changes and planning implications.

  \emph{Methods} For each urban agglomeration, four temporal snapshots from 1980 to 2016 have been derived from the land use inventory of the Swiss Federal Statistical Office.
  % A combination of landscape metrics, urban growth modes and fractal analysis serves to characterize the spatiotemporal patterns of urbanization.
  By making use of fractal analysis to study the area-radius relationship of urban land, each urban agglomeration is separated into two characteristic urban extents according to the distance of the city center, namely the inner and outer zones. The landscape metrics and growth modes are then computed at such extents. % and used to validate the diffusion and coalescence

  \emph{Results} % The fractal analysis of the area-radius relationship
  The time series of landscape metrics and growth modes reveal fairly different patterns when computed in the inner and outer zones respectively. Bern and Lausanne exhibit mostly traits of coalescence stages at the inner zone while displaying many characteristics of diffusion in the outer zone.
  % By contrast, Zurich, which is significantly more urbanized at higher distances to the city center and, shows
  In contrast, the trends of observed in the inner and outer zones of Zurich are both reminiscent of a coalescence of urban patches.

  \emph{Conclusions} Fractal analysis of the area-radius relationship can be a useful approach to detect characteristic extents of urban agglomerations at which distinct spatiotemporal patterns might be observed. Current models of urbanization patterns should incorporate the notion of characteristic extents more explicitly.

@@

% context
The last centuries have seen an unprecedented growth of urban areas, which has resulted in dramatic conversion of natural land and profound changes in landscape patterns and the ecosystem functions that they support \citep{alberti2005effects}.
The combination of current demographic prospects and the observed trends of decreasing urban densities suggest that the global amount of land occupied by cities might increase threefold by 2030 \citep{angel2005dynamics}.
Although the land use and land cover changes associated to urbanization have occurred on less than a 3\% of the earth's terrestrial surface, the environmental footprint of cities has significant implications at the global scale, for their functioning produces 78\% of the earth's greenhouse gases \cite{grimm2008global}.
% Quantifying urban landscape patterns in space and time is an important and necessary step to understand the driving forces and ecological impacts of urbanization \citep{wu2014urban}.
Given that urbanization will continue to be a major form of landscape change in the next decades, quantifying urban landscape patterns in space and time is crucial to understand the driving forces and ecological impacts of urbanization \citep{wu2014urban}.

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% % the morphology and evolution of cities
% % - location theory, gradient analysis, fractals: however, cities change in time, out-of-equilibrium
% % - spatiotemporal patterns and hypotheses
% % The study of urban morphology has been a central subject of interest for geographers and economists, especially following the spatial land-use model of \cite{von1826isolirte}.
% Understanding the spatial organization of cities is a subject of central interest in geography as well as other related disciplines.
% Following the seminal land-use model of \cite{von1826isolirte}, economists have developed a set of location theories with the aim of predicting the spatial distribution of economic activity and land use as a function of the distance to the city centers \citep{burgess1925growth,christaller1933zentralen,losch1954economics,alonso1964location}.
% % ``While many studies have found empirical correlations between environmental performance and aggregate measures of urbanization such as density or the extent of the built-up area, such measures do not reflect the complexity and diversity of existing urban patterns, and are strongly affected by the different definitions of city boundaries \cite{alberti1999urban}'' from \cite{bosch2019addressing}
% % Nevertheless, traditional density-based measures are unsuitable to describe the spatial complexity of the urban patterns observed in contemporary cities.
% Nevertheless, while traditional location theories are capable of predicting the land use changes and urban density decays as one moves away from the urban core, they fail to explain the complex patterns and fragmentation observed in the urban-rural fringe of contemporary cities \citep{irwin2002interacting}.
% % Besides density, measuring the complex spatial geometry of urban sprawl
% In response to such shortcoming, recent approaches to measure urban sprawl have built upon metrics from landscape ecology and fractal geometry \citep{frenkel2008measuring,torrens2008toolkit,bosch2019addressing}.
% % Paralleling the advances in location theory
% % On the other hand, during the last three decades, ecologists have largely built upon the gradient paradigm to evaluate how the environmental characteristics vary with decreasing human influence as one moves away from the urban cores \citep{mcdonnell1990ecosystem,medley1995forest,carreiro2005forest,mcdonnell2008ecosystem}.

% % TODO: intro to landscape ecology
% % The main focus of landscape ecology is the study of how the spatial pattern of landscapes affects the ecological processes that occur upon them \cite{turner1989landscape,pickett1995landscape}.
% The landscape ecology perspective views landscapes as an heterogeneous spatial mosaic of biological and physical patches, where the size, shape and configuration of patches exerts a strong influence on key ecosystem functions such as biodiversity and fluxes of organisms and materials \citep{turner1989landscape,pickett1995landscape}.
% In order to quantify such pattern-process relationships, a series of landscape metrics have been developed in order to the quantify the composition and configuration of landscapes and monitor their change through time \citep{o1988indices,turner1990spatial,mcgarigal2012fragstats}.
% % with large environmental footrprints
% From this standpoint, urban landscapes might be seen as patchy ecosystems, whose spatiotemporal evolution is strongly altered by human action \citep{pickett1997conceptual,collins2000new,zipperer2000application,wu2008making,mcphearson2016advancing}.
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Recent decades have witnessed an increasing number of studies of the spatiotemporal patterns of land use change associated to urbanization \citep{dietzel2005spatio, seto2005quantifying, schneider2008compact, jenerette2010global, wu2011quantifying, li2013quantifying, liu2016general, nong2018quantifying}.
Building upon previous ideas of urban growth phases and wave-like urban development, initial attempts to synthesis suggested that urbanization can be characterized as a two-step alternating process of diffusion and coalescence \citep{dietzel2005spatio, schneider2008compact}. Nonetheless, subsequent studies challanged the empirical validity of such hypothesis.
% On the one hand
The thorough study of \cite{jenerette2010global} examined the spatiotemporal patterns of land use change of a sample of 120 cities distributed throughout the world from 1990 to 2000, and determined that overall, urbanization leads to fragmented landscapes with more complex and heterogeneous structures.
Similarly, in a comparative analysis of the metropolitan regions of Phoenix and Las Vegas, \cite{wu2011quantifying} revealed that throughout the 20th century, the two agglomerations did not display signs of distinct urban growth phases, but instead showed a strikingly similar trend towards a landscape that is more diverse in land use, fragmented in structure and complex in shape.
% On the other hand
Subsequently, \cite{li2013quantifying} determined that the two-phase diffusion and coalescence model can be over-simplistic and that urbanization might be better characterized by means of three growth modes, namely infilling, edge expansion and leapfrogging, which operate simoultaneously while alternating their relative dominance.
Such results were confirmed by the thorough study of 16 world cities over the 1800-2000 period by \cite{liu2016general}, who further resolved that urbanization generally leads to an increasingly diverse and complex landscape.

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% % TODO: introduce the hierarchy idea: ``City morphology is reflected in a hierarchy of different subcenters or clusters accross many scales, from the entire city to neighborhoods, organized around key economic functions'' \cite{batty2008size}. Also \cite{wu1995balance, liu2014much}.
% % ``Because urban systems are multi-scaled, social-ecological systems, a hierarchical approach is needed for understanding their structure, function and dynamics'' \cite{li2013quantifying}. See also \cite{pickett1997conceptual}
% % TODO: 
% % TODO: maybe remove the ``classic location theory'' part. But note \cite{white2015modeling}: despite complexity, cities are configured by concentric rings. Also note in discussion that as Swiss agglomerations fuse, the concentric model is very likely to "not hold" and a poly-centric (e.g., multi-fractal) model must be considered.
% Understanding the spatial organization of cities is a subject of central interest in geography as well as other related disciplines.
% % Following the seminal land-use model of \cite{von1826isolirte}, economists have developed a set of location theories with the aim of predicting the spatial distribution of economic activity and land use as a function of the distance to the city centers \citep{burgess1925growth,christaller1933zentralen,losch1954economics,alonso1964location}.
% Following the seminal land-use model of \cite{von1826isolirte}, economists have developed a set of location theories with the aim of predicting the spatial distribution of economic activity and land use in human settlements \citep{burgess1925growth,christaller1933zentralen,losch1954economics,alonso1964location}.
% % However, while traditional location theories are capable of predicting the land use changes and urban density decays as one moves away from the urban core, they fail to explain the complex patterns and fragmentation observed in the urban-rural fringe of contemporary cities \citep{irwin2002interacting}.
% However, while traditional location theories are capable of reproducing the land use changes and urban density decays as one moves away from the urban core, they fail to explain the spatial complexity encountered in contemporary cities.
% % Despite apparent complexity and diversity
% % Paralleling the above studies, approaches from the complexity sciences have provided novel insights into the spatial organization that underpins contemporary cities \citep{batty2005cities}.
% In response to such shortcomings, during the last decades the study of urban morphology has increasingly borrowed from concepts from the complexity sciences \citep{batty2005cities, white2015modeling}.
% % Like many other complex systems, cities exhibit morphological traits that are consistent with fractal geometry and reflect the self-organizing nature of the processes that occur upon them \citep{white1993cellular}.
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% TODO: Current hypothesis of spatiotemporal patterns of urbanization either overlook that cities show distinct spatial patterns depending on the distance to the city center (diffusion and coalescence), or consider it by making use of artificial extents, i.e., administrative boundaries (three growth modes).
% administrative boundaries are ``constructed  manually case-by-case based on subjective judgement'' \cite[page 18702]{rozenfeld2008laws}
% Nevertheless, such spatiotemporal models of urbanization missappreciate the way in which cities show distinct spatial signatures along the urban-rural gradient. On the one hand, such notion is entirely overlooked in the formulation of the diffusion and coalescence model by \cite{dietzel2005spatio}. On the other hand, although \cite{li2013quantifying} employed a hierarchical framework of three nested spatial extents in order to test their three growth modes hypothesis, their choice of extents is based on administrative boundaries, which are constructed manually case-by-case and often based on subjective judgement and might thus lead to equivocal results \citep{wolman2005fundamental,rozenfeld2008laws, liu2014much, oliveira2014large}.
Nevertheless, such models of urbanization missappreciate the way in which contemporary cities are multi-scaled systems, organized in different levels that show its own characteristic spatiotemporal patterns \citep{batty2005cities, batty2008size, white2015modeling}.
While both the diffusion and coalescence model of \cite{dietzel2005spatio} and the three growth modes model of \cite{li2013quantifying} make use of a hierarchical framework and evaluate the spatial patterns at different extents, the choice of such extents is not based on quantitative criteria and neglects the characteristic scales of complex systems such as urban patterns.
Despite the apparent complexity and diversity of urban forms, cities comply with well-defined principles of spatial organization, which can be characterized quantitatively by means of fractal geometry.
% Although the scaling relationships of fractal geometry suggest the existence of strong morphological regularities over a wide variety of cities and regions \citep{frankhauser1994fractalite, batty1994fractal, white2015modeling}, their meaning in the context of the spatiotemporal patterns of urbanization remains unclear \citep{li2000fractal, manson2006complexity, bosch2019addressing}.
% More precisely, the relationship between the total built-up area and the distance from the city center has been shown to empirically follow a scaling relationship with very stable exponents for a wide variety of cities \citep{frankhauser1994fractalite, batty1994fractal}.
A remarkable regularity is found in the relationship between the total built-up area and the distance from the city center, which has been shown to empirically follow a scaling law with very stable exponents for a wide variety of cities \citep{frankhauser1994fractalite, batty1994fractal}.
% Like classic location theory, fractal geometry models how space is filled as one moves away from the city center, a characteristic that can hardly be discerned by landscape metrics and growth modes. Yet unlike traditional density curves, fractal geometry is also capable of reproducing the complex and fragmented shapes encountered at the urban-rural fringe of contemporary cities.
% Further examination by \cite{white1993cellular} suggested that the area-radius scaling could be better approximated throguh two scaling exponents, a first steeper one for small distances to the city center, reflecting an inner zone where urbanization was essentially complete, and a second lower slope for the outer zone that is still undergoing urbanization.
In a thorough examination of a global sample of cities, \cite{frankhauser1994fractalite} noted the existence of a kink in the area-radius relationship, which reveals a change on the spatial structure of cities at a certain distance from their center.
The same pattern was found in the urban cellular automata simulations of \cite{white1993cellular}, suggesting that the area-radius scaling could be better approximated throguh two scaling exponents, a first steeper one for small distances to the city center, reflecting an inner zone where urbanization was essentially complete, and a second lower slope for the outer zone that is still undergoing urbanization.

The objective of this study is therefore to build upon fractal analysis in order to enlighten the current hypotheses of the spatiotemporal patterns of urbanization.
More precisely, the area-radius relationship will be used to detect characteristic extents in urban agglomerations, such as the inner and outer zones reviewed above.
Thereupon, the time series of landscape metrics and growth modes will be computed at such extents in order to evaluate the degree to which the spatiotemporal patterns of urbanization operate differently at each scale.
The results will serve assess the validity of the diffusion and coalescence and three growth modes hypothesis and provide critical insights into how they could be revised from a multi-scale perspective.

@@

Given that a significant part of the cross-border urban agglomerations of Geneva and Basel (the second and third largest in Switzerland) lie beyond the Swiss boundaries \mycitep{sfso2014espace}, in order to ensure coherence of the land use/land cover data, this study comprises only three of the five largest Swiss urban agglomerations, namely Bern, Lausanne and Zurich \mycitep{sfso2018city}.
As shown in \autoref{figures/population_change.pdf}, the three agglomerations have undergone important population growth over the last 30 years, especially during the most recent years and at the agglomeration extent.
With a total population over 1.3 million and land area of 1305 km$^2$ (1038 hab/km$^2$), Zurich is the largest Swiss urban agglomeration. As a leading global city and one of the world's largest financial centers, Zurich has the country's largest airport and railway station, and also hosts the largest Swiss universities and higher education institutions. 
Bern is the capital of Switzerland and fourth most populous urban agglomeration in Switzerland, with a total population of 410000 inhabitants and occupying a land area of 783 km$^2$ (531 hab/km$^2$). As the fifth largest Swiss urban agglomeration and the second most important student and research center after Zurich, the Lausanne agglomeration has a total population of 409000 inhabitants over a land area of 773 km$^2$ (537 hab/km$^2$). Given its larger population growth rate, Lausanne is likely to soon surpass Bern and become the fourth largest urban agglomeration in Switzerland.
% Zurich: 1021859 (1990), 1080728 (2000), 1191058 (2010), 1354140 (2017)
% Bern: 351084 (1990), 349157 (1990), 357668 (2010), 415785 (2017)
% Lausanne: 300280 (1990), 311441 (2000), 339,389 (2010), 415596 (2017)
% Densities: Zurich: 1037 Bern: 531, Lausanne: 537
Overall, the three urban agglomerations are characterized by a pervasive public transportation system and a highly developed economy, with a 85\% of the employment devoted to the tertiary sector.

@@

\begin{figure}[!ht]
  \begin{adjustwidth}{0cm}{-.4\textwidth}
    \centering  
    \includegraphics[width=\linewidth]{figures/landscape_plots.pdf}
    % \vspace{.5em}
    \caption[Evolution of urban patches]{\label{figures/landscape_plots.pdf}Evolution of urban patches of the three urban agglomerations throughout their respective periods of study. The times t$_0$, t$_1$, t$_2$ and t$_3$ correspond to 1981, 1993, 2004 and 2013 for Bern; 1980, 1990, 2005 and 2014 for Lausanne and 1982, 1994, 2007 and 2016 for Zurich.}
  \end{adjustwidth}
\end{figure}

% \subsection*{Fractal aspects of urban patterns}
\subsection*{Area-radius scaling in urban agglomerations}

% Despite their complex and irregular appearance, cities comply with well-defined order principles that can be characterized quantitatively by means of fractal geometry \citep{frankhauser1994fractalite, batty1994fractal}.
% Two main characteristics of fractal structures are of particular interest in the context of the study of urbanization and land use change \citep{white2015modeling}: the area-radius scaling and the size-frequency distribution of urban patches.
% In order to evaluate how each agglomeration fills the available space, fractal geometry will be employed to study the relationship between the built-up area of an urban agglomeration and the distance from the main city center.
In order to quantitatively detect characteristic spatial extents of urban agglomerations, the relationship between the built-up area and the distance form the main city center will be evaluated from the perspective of fractal geometry.
% \paragraph*{Area-radius scaling} \hspace{0cm} % force line break
% The relationship between the built-up area of an urban agglomeration and the distance from the main city center has been shown (e.g., \cite{frankhauser1994fractalite}) to empirically conform to relationships of the form:
% More precisely, it has been shown, e.g., by \cite{frankhauser1994fractalite}, that such relationship can be empirically approximated as in:
% For fractal objects, such relationship should follow a scaling rule of the form:
If cities are to be considered fractal objects, such relationship should follow a scaling rule of the form \citep{mandelbrot1983fractal}:

where $A$ denotes the total area of the urban built-up extent that lays within a distance $r$ from the city center, and $D$ corresponds to the radial dimension, analogous to the fractal dimension of two-dimensional complex objects such as Sierpinski carpets.
% Although the measure might not be appropriate for urban agglomerations with multiple important centers, \cite{frankhauser1994fractalite} found extensive evidence that most contemporary cities could be approximated through (\ref{eq:radial-dimension}), with the value of $D$ consistently falling between 1.9 and 2.0.

% On the other hand, following initial observations by \cite{frankhauser1990aspects}, \cite{white1993cellular} suggested that the area-radius scaling could be better approximated throguh two values of $D$, a first steeper one for small values of $r$, reflecting an inner zone where urbanization was essentially complete, and a second lower slope for the outer zone that is still undergoing urbanization.
% Therefore, in order to evaluate whether the area-radius relationship should be fit by one or two curves, the coefficient of determination of a piecewise linear regression with two segments will be compared to that of a simple linear regression. The optimal breakpoint of the two-segment regression, namely, the breakpoint location that minimizes the sum of squared residual will be computed with the \texttt{pwlf} Python library\footnote{See \url{https://github.com/cjekel/piecewise_linear_fit_py}}, which is based on the Differential Evolution (DE) optimization algorithm \citep{storn1997differential}.
With the aim of assessing whether the urban agglomerations follow the bi-fractal city model suggested by \cite{white1993cellular}, a piecewise linear regression with two segments will be compared to that of a simple linear regression. The optimal breakpoint of the two-segment regression, namely, the breakpoint location that minimizes the sum of squared residual will be computed with the \texttt{pwlf} Python library\footnote{See \url{https://github.com/cjekel/piecewise_linear_fit_py}}, which is based on the differential evolution optimization algorithm \citep{storn1997differential}.
In this context, such breakpoint corresponds to the kink in the area-radius scaling noted by \cite{frankhauser1994fractalite}, namely the radial distance to the city center at which cities show a distinct spatial structure that is less space-filling.
% In order to evaluate the signatures that the spatiotemporal patterns of urbanization show at different spatial extents, the metrics will be computed at three spatial extents. The first correspond to the whole urban agglomeration, as defined by the SFSO \mycitep{sfso2014espace}. Following the fractal analysis of the area-radius scaling, the inner zone, defined as the
% In consonance with the multi-scale perspective
Thereupon, three spatial extents will be considered in the analysis of the spatiotemporal patterns of urbanization. The first extent corresponds to the whole urban agglomeration defined by the SFSO \mycitep{sfso2014espace}, which is described in the foregoing section. The second and third extents will be derived from the location of the kink, i.e., the breakpoint of the two-segment regression of the area-radius relationship. More precisely, in line with \cite{white1993cellular}, the second extent will be defined as the inner zone, i.e., a circle with with the city core as center and the breakpoint distance as radius, while the third extent will be defined as the outer zone, i.e., the area that lies outside the inner zone circle and the agglomeration boundaries.

% \paragraph*{Size-frequency distribution of urban patches} \hspace{0cm} % force line break

% Contemporary urban agglomerations are configured by numerous patches of urban land use.
% If such configuration is fractal, there must be no characteristic patch size, and thus the relationship between the size of an urban patch and the number of patches of that size found in the agglomeration must follow a power-law scaling relationship of the form:

% \begin{equation}
%   \label{eq:size-frequency}
%   N(s) \sim s^{-\alpha}
% \end{equation}

% where $N(s)$ is the number of patches of size $s$, and the scaling exponent $\alpha$ is the patch size-frequency dimension \citep{white1993cellular}.

@@

\subsubsection*{Time series of landscape metrics}

While a plentiful collection of landscape metrics can be found in the literature, many of them are highly correlated with one another. As a matter of fact, \cite{riitters1995factor} found that the characteristics discerned by 55 prevalent landscape metrics could be reduced to only 6 independent factors.
On the other hand, landscape metrics can be very sensitive to the resolution and the extent of the maps. However, several metrics empirically exhibit consistent responses to changing scales that conform to predictable scaling relations \citep{wu2002empirical, wu2004effects}.
Based on such remarks, and in order to enhance comparability with other studies, ten landscape metrics have been selected for the present study, whose details are listed in \autoref{selected-metrics}.

\begin{table}[!h]
  % \begin{adjustwidth}{-.4\textwidth}{0cm} % Comment out/remove adjustwidth environment if table fits in text column.
  \footnotesize % Font size can be changed to match table content. Recommend 10 pt.
  \renewcommand{\arraystretch}{1.5} % vertical spacing
  \centering
  \caption[Selected landscape metrics]{\label{selected-metrics}Selected landscape metrics to quantify the spatiotemporal patterns of urbanization. A more thorough description can be found in the documentation of the software FRAGSTATS v4 \citep{mcgarigal2012fragstats}}
  % \begin{tabular}{p{.3\textwidth} p{.22\textwidth} p{.16\textwidth} p{.6\textwidth}} 
  %   \toprule
  %   \textbf{Metric name (abbrev.)} & \textbf{Level} & \textbf{Category} & \textbf{Description} \\
  %   \midrule
  %   %   Percentage of landscape (PLAND) & Class & Area and edge & Percentage of landscape, in terms of area, occupied by patches of a given class \\
  %   %   Mean patch size (MPS) & Class and landscape & Area and edge & Mean patch size \\
  %   %   Largest patch index (LPI) & Class and landscape & Area and edge & Percentage of the landscape occupied by the largest patch \\
  %   Patch density (PD) & Class and landscape & Aggregation & The number of patches per area unit \\
  %   Edge density (ED) & Class and landscape & Area and edge & Sum of the lengths of all edge segments per area unit \\
  %   Area-weighted mean fractal dimension (AWMFD) & Class and landscape & Shape & Mean patch fractal dimension weighted by relative patch area \\
  %   Mean euclidean nearest neighbor distance (ENN) & Class and landscape & Aggregation & Mean patch shortest edge-to-edge distance to the nearest neighboring patch of the same or different class \\
  %   %   Landscape shape index (LSI) & Class and landscape & Aggregation & Standardized ratio of edge length to area \\
  %   %   Contagion (CONTAG) & Landscape & Aggregation & Metric that measures the extent to which patches of the same class are spatially aggregated in the landscape \\
  %   %   Shannon's diversity index (SHDI) & Landscape & Diversity & Metric that measures the diversity of patch types determined by both the number of different patch types and the proportional distribution of area among patch classes. \\
  %   \bottomrule  
  % \end{tabular}
  \begin{tabular}{p{.28\textwidth} p{.16\textwidth} p{.46\textwidth}} 
    \toprule
    \textbf{Metric name} & \textbf{Category} & \textbf{Description} \\
    \midrule
    Percentage of landscape (PLAND) & Area and edge & Percentage of landscape, in terms of area, occupied by patches of a given class \\
    Patch density (PD) & Aggregation & The number of patches per area unit \\
    Edge density (ED) & Area and edge & Sum of the lengths of all edge segments per area unit \\
    Area-weighted mean fractal dimension (AWMFD) & Shape & Mean patch fractal dimension weighted by relative patch area \\
    Mean euclidean nearest neighbor distance (ENN) & Aggregation & Mean patch shortest edge-to-edge distance to the nearest neighboring patch of the same or different class \\
    \bottomrule  
  \end{tabular}    
  % \end{adjustwidth}
\end{table}

While complying with the FRAGSTATS v4 definitions \citep{mcgarigal2012fragstats}, the landscape metrics have been computed with the open source library PyLandStats \citep{bosch2019pylandstats}.
Like in most of the related studies, the categorical maps have been reclassified into urban and natural classes, and the metrics have computed at the urban class level, namely aggregating their values across all the urban patches of the landscape.
% The only exceptions are the contagion (CONTAG) and Shannon's diversity index (SHDI), which can only be computed at the landscape level, namely considering all the landscape classes (urban and natural in the this study).
% The only exception is the contagion index (CONTAG), which can only be computed at the landscape level, namely considering all the landscape classes (urban and natural in the this study).
Pixels that correspond to land unavailable for development, such as water bodies, have been excluded from the computation of the metrics.
